---
title: "FA1-DMW-BAYQUEN"
output: html_document
date: "2024-02-16"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Lecture 2: Data Visualization

## **VARIATION**

**Discrete Variables**

```{r}
library('ggplot2')
ggplot(data = diamonds) +
  geom_bar(mapping = aes(x = cut), fill="salmon")
```

**Continuous Variables**
```{r}
ggplot(data = diamonds) +
geom_histogram(mapping = aes(x = carat), fill="salmon")
```

> *zoomed in portion of the plot using coord_cartesian:*

```{r}
ggplot(data = diamonds) +
  geom_histogram(mapping = aes(x = carat),fill="salmon") +
  coord_cartesian(xlim = c(0,3))
``` 

> *decreasing the bandwith*

```{r}
ggplot(data = diamonds) +
  geom_histogram(mapping = aes(x = carat), binwidth = 0.01, fill = "salmon") +
  coord_cartesian(xlim = c(0,3))
```

## **COVARIATION**

**Discrete versus discrete**

```{r}
ggplot(data = diamonds,) +
  geom_count(mapping = aes(x = cut, y = clarity), color="darkseagreen3")

```

**Continuous versus continuous**

> **Many to one**

```{r}
ggplot(data = diamonds) +
  geom_point(mapping = aes(x = x, y = y), color="darkseagreen3")
```

> *zoomed in on the part of the plot where most of the points lie*

```{r}
ggplot(data = diamonds) +
  geom_point(mapping = aes(x = x, y = y), color="darkseagreen3") +
  coord_cartesian(xlim = c(3.5, 10), ylim = c(3.5,10))
```

> *add a line to the plot via geom_abline()*

```{r}
ggplot(data = diamonds) +
  geom_point(mapping = aes(x = x, y = y),  color="darkseagreen3") +
  geom_abline(slope = 1, intercept = 0, color = "red") +
  coord_cartesian(xlim = c(3.5, 10), ylim = c(3.5,10))
```

> **One to one**

construct the requisite data frame

```{r}
library(dplyr)

avg_price_by_carat <- diamonds %>%
  mutate(carat = round(carat, 1)) %>%
  group_by(carat) %>%
  summarise(avg_price = mean(price))
avg_price_by_carat
```

> *line plot*


```{r}
ggplot(data = avg_price_by_carat) +
  geom_line(mapping = aes(x = carat, y = avg_price))
```

**Continuous versus discrete**

> *boxplots*

```{r}
ggplot(data = diamonds) +
  geom_boxplot(mapping = aes(x = cut, y = price))
```

## **ADDITIONAL VISUALIZATION TOOLS**

> **Axis Transformations**

```{r}
ggplot(data = diamonds) +
  geom_histogram(aes(x = carat), binwidth = 0.01, color = "darkorchid2")

```

> *x-axis on a logarithmic scale*

```{r}
ggplot(data = diamonds) +
  geom_histogram(aes(x = carat), binwidth = 0.01, color = "darkorchid2") +
  scale_x_log10()

```
 
> *y-axis on a logarithmic scale*

```{r}
ggplot(data = diamonds) +
  geom_point(aes(x = carat, y = price), color = "darkorchid2") +
  scale_x_log10() +
  scale_y_log10()
```
 
> **Multiple aesthetics in the same geom**

```{r}
ggplot(data = diamonds) +
  geom_point(mapping = aes(x = carat, y = price))
```

> *color points based on the cut of the diamond*

```{r}
ggplot(data = diamonds) +
  geom_point(mapping = aes(x = carat, y = price, color = cut))
```

> *Faceting*

```{r}
ggplot(data = diamonds) +
  geom_point(mapping = aes(x = carat, y = price, color = cut)) +
  facet_wrap(~cut)
```

```{r}
ggplot(data = diamonds) +
  geom_point(mapping = aes(x = carat, y = price, color = cut)) +
  facet_grid(. ~ cut)
```
```{r}

ggplot(data = diamonds) +
  geom_point(mapping = aes(x = carat, y = price, color = cut)) +
  facet_grid(clarity ~ cut)
```
