---
title: "FA1-2-DMW-BAYQUEN"
output: html_document
date: "2024-02-16"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Lecture 3: Data Transformation
```{r}
library(tidyverse)
```

## **ISOLATING DATA**
> using the **filter()** function

*subset to diamonds with price at least $10,000*

```{r}
filter(diamonds, price >= 10000)
```

*subset to diamonds with price at least $10,000 AND clarity VVS1 or IF* 

```{r}
filter(diamonds, price >= 10000 & clarity %in% c("VVS1", "IF"))
```

> using the **select()** function

*select columns corresponding to the "4 C's"*
```{r}
select(diamonds, carat, cut, color, clarity)
```

> using the **arrange()** function

*sort diamonds by carat (ascending)*
```{r}
arrange(diamonds, carat)
```

*sort diamonds by carat (descending)*
```{r}
arrange(diamonds, desc(carat))
```

## **DERIVING INFORMATION**

> using the  **mutate()** funciton

*add column that is the price per carat of each diamond*
```{r}
mutate(diamonds, price_per_carat = price/carat)
```
*add column that indicates whether a diamond's price per carat is at least $10k*
```{r}
mutate(diamonds, fancy_diamond = price/carat > 10000)
```

*use if_else() if you have two cases*
```{r}
mutate(diamonds,
  good_value =
    if_else(
      condition = carat > 2, # check whether carat > 2
      true = price < 5000, # if so, good value if cheaper than $5k
      false = price < 1000 # if not, good value if cheaper than $1k
    )
)
```

*use case_when() if you have more than two cases*
```{r}
mutate(diamonds,
  value = case_when(
    carat > 2 & price < 5000 ~ "good", # if carat > 2 and price < 5000, then good
    carat > 1 & price < 2500 ~ "ok", # if carat > 1 and price < 2500, then ok
    TRUE ~ "bad" # otherwise, bad
  )
)
```


> using the  **sumarise()** funciton

*find the number of "fancy" diamonds (price per carat at least $10000)*
```{r}
summarise(diamonds, num_fancy_diamonds = sum(price/carat > 10000))
```

*find the number of "fancy" diamonds (price per carat at least $10000),as well as the mean price of a diamond*
```{r}
summarise(diamonds,num_fancy_diamonds = sum(price/carat > 10000),mean_diamond_price = mean(price))
```
## **MULTISTEP TRANSFORMATIONS**
> using **The pipe (%>%)**

```{r}
library(dplyr)

diamonds %>% # pipe in the data
  filter(cut == "Premium") %>% # restrict to premium cut diamonds
  mutate(price_per_carat = price/carat) %>% # add price_per_carat variable
  arrange(desc(price_per_carat)) # sort based on price_per_carat
```
passing data between different tidyverse packages, e.g. from dplyr to ggplot2
```{r}
diamonds %>% # pipe in the data
  filter(cut == "Premium") %>% # restrict to premium cut diamonds
  mutate(price_per_carat = price/carat) %>% # add price_per_carat variable
    ggplot() + # start a ggplot
    geom_histogram(aes(x = price_per_carat), fill="salmon") # add a histogram
```

>using **group_by()**

```{r}
diamonds %>% # pipe in the data
  group_by(cut) %>% # group by cut
  summarise(max_price = max(price)) # find the max price for each cut
```
```{r}
diamonds %>% # pipe in the data
  group_by(cut, clarity) %>% # group by both cut and clarity
  summarise(max_price = max(price)) # find the max price for each group
```
>with ungrouping

```{r}
diamonds %>% # pipe in the data
  group_by(cut, clarity) %>% # group by both cut and clarity
  summarise(max_price = max(price)) %>% # find the max price for each group
  ungroup() # remove grouping
```

> **Storing the Transformed Data**

```{r}
max_prices <- diamonds %>% # pipe in the data
  group_by(cut) %>% # group by cut
  summarise(max_price = max(price)) # find the max price for each cut

max_prices
```