---
title: "FA1-3-DMW-BAYQUEN"
output: html_document
date: "2024-02-16"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Lecture 4:  Data Wrangling

```{r cars}
library(tidyverse)
heights <- read_csv("C:/Users/chris/Desktop/Productivity/Academics/College/3rd Year/2nd Semester/DATA MINING AND WRANGLING/FA1/heights.csv")
heights
```


```{r}
table1
table2
table3
table4a
table4b
```

## **PIVOTING**
> using the **pivot_longer()** function

```{r}
table4a %>%
pivot_longer(cols = c(`1999`, `2000`), names_to = "year", values_to = "cases")
```
Note that 1999 and 2000 are non-syntactic names (because they don’t start with a letter) so we have to
surround them in backticks. In the final result, the pivoted columns are dropped, and we get new year and cases columns. Otherwise, the relationships between the original variables are preserve

>using the **pivot_wider()** function; when an observation is scattered across
multiple rows.


```{r}
table2 %>%
pivot_wider(names_from = type, values_from = count)
```

## **JOINING**
recalling the previous tables
```{r}
tidy4a <- table4a %>%
pivot_longer(c(`1999`, `2000`), names_to = "year", values_to = "cases")
tidy4b <- table4b %>%
pivot_longer(c(`1999`, `2000`), names_to = "year", values_to = "population")
tidy4a
```

applying the **left_join()** function to tidy4a and tidy4b

```{r}
left_join(x = tidy4a, y = tidy4b, by = c("country", "year"))
```

## **SEPARATING**

pulls apart one column into multiple columns, by splitting whenever a separator character
appears

```{r}
table3
table3 %>%
separate(rate, into = c("cases", "population"), sep = "/", convert = TRUE)
```

## **MISSING VALUES**
Missing values, marked with NA, are often present in real datasets.

```{r}


stocks <- tibble(
  year = c(2015, 2015, 2015, 2015, 2016, 2016, 2016),
  qtr = c( 1, 2, 3, 4, 2, 3, 4),
  return = c(1.88, 0.59, 0.35, NA, 0.92, 0.17, 2.66)
  )
stocks
```
pivoting wider
```{r}
stocks %>%
pivot_wider(names_from = year, values_from = return)

```

The simplest approach to dealing with missing values in a dataset is to remove all
rows containing any missing values. This can be done via **na.omit()**. 

```{r}
stocks %>%
  na.omit()
```